Pattern Avoidance

### 2014-11-07 project2/paper/paper.tex
added more rigor to the proof to Theorem 28

\begin{proof}
Let $A$ be the set of permutations in $T_n$ avoiding 312, and $B$ the set of permutations in $T_n$ avoiding 321. We will prove the theorem by constructing a bijection between $A$ and $B$.

Let $a$ be an arbitrary permutation in $A$, and $b$ an arbitrary permutation in $B$. By the definition of $A$, $a$ avoids 312. Pick any positive integer $i\leq n$. We note that all even integers to the right of $i$ and less than $i$ in $a$ must be in decreasing order; otherwise, we would obtain a subsequence with pattern 312. Similarly, all even integers to the right of $i$ and less than $i$ in $b$ must be in increasing order.

Let $a$ be an arbitrary permutation in $A$. Clearly, by the definition of $A$, $a$ must avoid $312$. Now, we claim that for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in decreasing order; if this were not the case, then we would have a subsequence in the relative ordering $312$ (the element $i$ followed  by the two increasing even numbers less than $i$), which would be a contradiction. It is easy to see that we can make a corresponding claim about permutation $b$, that is for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in increasing order.

The above two claims show that there is actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$. To obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order). Similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order). Note that these transformations can be expressed as multiple swaps (as described in Lemma \ref{swapping_lemma}). \todo{should have some justification here that reversing reordering these elements doesn't 'step on each other'... that is, you can always do it for some $i_1$ and then some $i_2$ afterward}
Since the mapping from $A$ to $B$ is clearly reversible, and hence invertible, we conclude that the mapping we just established between $A$ and $B$ is bijective.

Since there exists a bijective mapping between permutations in $T_n$ that avoid $312$ and permutations in $T_n$ that avoid $321$, we can conclude that $t_n(312) = t_n(321)$.

>>>>

\begin{proof}
Let $A$ be the set of permutations in $T_n$ avoiding 312, and $B$ the set of permutations in $T_n$ avoiding 321. Let $a$ be an arbitrary permutation in $A$, and $b$ an arbitrary permutation in $B$. By the definition of $A$, $a$ avoids 312. Pick any positive integer $i\leq n$. We note that all even integers to the right of $i$ and less than $i$ in $a$ must be in decreasing order; otherwise, we would obtain a subsequence with pattern 312. Similarly, all even integers to the right of $i$ and less than $i$ in $b$ must be in increasing order.

In fact, note that any permutation $\pi\in T_n$ which satisfies this condition - for all $i\leq n$, all the even integers to the right of $i$ and less than $i$ are in decreasing order - is in fact a permutation in $A$. Suppose for the sake of contradiction $\pi$ had a subsequence $x, y, z$ following the pattern 312 (so in particular $x>z>y$). Since it is in $T_n$, by Lemma  \ref{tn_avoidance} $y$ and $z$ must be even. Then by the assumed property of $\pi$ we must have $y>z$, a contradiction; hence such a subsequence cannot exist, and so $\pi$ avoids 312. We have a similar result for $B$: $B$ is precisely the set of permutations $\pi\in T_n$ satisfying the condition that, for all $i\leq n$, all even integers to the right of $i$ and less than $i$ are in decreasing order.

Then the bijection between $A$ and $B$ is as follows: given a permutation $a\in A$, consider each of its elements $i$ beginning with its rightmost element and moving leftward. By the argument above, the collection of even elements to the right of $i$ and less than $i$ will be in decreasing order; for each $i$, we sort this collection to make the elements increasing instead. Note that this process can be obtained by transposing elements, so that by Lemma \ref{swapping_lemma} the result at each stage will be a permutation in $T_n$. The end result will therefore be a permutation in $T_n$ with the property that for each $i$, all even integers to the right of $i$ and less than $i$ must be in increasing order. As we argued above, therefore, the end result is a permutation in $B$.

Indeed, it is easy to see that this process is invertible: for each element $j$, beginning with the leftmost element and moving right, we reverse sort the elements less than $j$ and to the right of $j$. Since the process is invertible, this map in fact defines a bijection, as desired.
\end{proof}
